---
title: "Forward TS recorder workflow and bridge comparison"
output: html_document
date: "2026-07-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## General Notes

This notebook shows the current forward tree-sequence recording workflow.
It uses the same forward simulation pattern as `dev/notesRealBreakpoints.md`:

1. create founders from tree sequences,
2. run a short AlphaSimR breeding loop,
3. record meiosis breakpoints in forward time,
4. compare the direct forward TS recorder with the older
   `bridgeCollectSegGenFromSimOutput()` path,
5. write `.trees` files that can be compared in Python.

Two founder entry points are shown:

1. start from `runMacsTS(...)`,
2. start from existing msprime `.trees` files.

### Main workflow differences between old "bridge" and forward ts recorder:

| Aspect | Old bridge path | Current forward recorder |
|---|---|---|
| When TS edges are made | After simulation | During each cross |
| Required recombination record | Stored `SP$recHistGen` | Temporary `tsSegGen` matrix only |
| Main stored object during simulation | `SP$recHistGen` | `attr(SP, "tsForwardRecorder")` |
| Segment data kept by default | Reconstructed later for bridge output | Not kept by default |
| Role of `SP$pedigree` | Used with `recHistGen` to reconstruct child-parent edge relationships after simulation | Still used for TS individual parent links, node times, and default sample inference, but not as the main source of inheritance edges |
| Recombination map source | Supplied in validation `chr_info` when bridge converts `recHistGen` to coordinates | Stored on the founder as `tsForwardPosMeta`; from `runMacsTS()` defaults/`-R` or explicit `asMapPop(chr_info)` maps |
| User output function | `bridgeWriteTrees()` | `tsForwardFinalizeFromSimParam()` |
| Metadata timing | Added when bridge writes trees | Added while recording nodes/individuals |
| Edge append path | R row loop | C++ bulk append |
| Child node append path | R row loop | C++ bulk append, with or without metadata |

In short, the bridge path stores recombination history first and reconstructs
tree-sequence rows later. The forward recorder uses transient meiosis segments
to append tree-sequence rows immediately. It still uses pedigree information
for TS individual rows and node timing, but normal use does not need to keep
`recHistGen` or `tsSegGen`.

## 1) Setup

```{r libraries}
library(RcppTskit)
library(jsonlite)
library(ggplot2)
library(AlphaSimR)
devtools::load_all()

out_dir <- 'testData'
```

## 2) Validation Helpers

These helpers are only for interpretation and validation. The actual workflow
steps are kept in the main sections below so the user-facing process is clear.

```{r helpers}
normalize_edge_df <- function(df) {
  out <- data.frame(
    chr = as.integer(df$chr),
    child = as.integer(df$child),
    parent = as.integer(df$parent),
    left = as.numeric(df$left),
    right = as.numeric(df$right),
    stringsAsFactors = FALSE
  )
  out <- out[out$right > out$left, , drop = FALSE]
  if (nrow(out) == 0L) {
    return(out)
  }
  ord <- order(out$chr, out$child, out$parent, out$left, out$right)
  out <- out[ord, , drop = FALSE]
  rownames(out) <- NULL
  out
}

collect_forward_edges_for_children <- function(recorder, chr, child_iids) {
  chrState <- recorder$chr[[chr]]
  tc <- chrState$tc
  n_edge <- as.integer(tc$num_edges())
  out <- vector("list", n_edge)
  k <- 0L

  iid_keys <- ls(chrState$indMap, all.names = TRUE)
  if (length(iid_keys) == 0L) {
    return(data.frame(chr = integer(), child = integer(), parent = integer(),
                      left = numeric(), right = numeric(), stringsAsFactors = FALSE))
  }
  ind_rows <- vapply(iid_keys, function(key) {
    as.integer(get(key, envir = chrState$indMap, inherits = FALSE))
  }, integer(1))
  ind_row_to_iid <- setNames(as.integer(iid_keys), as.character(ind_rows))

  for (i in seq_len(n_edge)) {
    e <- tc$edge_table_get_row(i - 1L)
    child_node <- as.integer(e$child)
    parent_node <- as.integer(e$parent)
    child_ind_row <- as.integer(tc$node_table_get_row(child_node)$individual)
    parent_ind_row <- as.integer(tc$node_table_get_row(parent_node)$individual)

    child_iid <- ind_row_to_iid[as.character(child_ind_row)]
    parent_iid <- ind_row_to_iid[as.character(parent_ind_row)]
    if (length(child_iid) == 0L || length(parent_iid) == 0L ||
        is.na(child_iid) || is.na(parent_iid)) {
      next
    }
    child_iid <- as.integer(child_iid)
    if (!(child_iid %in% child_iids)) {
      next
    }

    k <- k + 1L
    out[[k]] <- data.frame(
      chr = as.integer(chr),
      child = child_iid,
      parent = as.integer(parent_iid),
      left = as.numeric(e$left),
      right = as.numeric(e$right),
      stringsAsFactors = FALSE
    )
  }

  if (k == 0L) {
    return(data.frame(chr = integer(), child = integer(), parent = integer(),
                      left = numeric(), right = numeric(), stringsAsFactors = FALSE))
  }
  do.call(rbind, out[seq_len(k)])
}

collect_forward_edges_all_chr <- function(recorder, child_iids) {
  out <- lapply(seq_len(recorder$nChr), function(cc) {
    collect_forward_edges_for_children(recorder, chr = cc, child_iids = child_iids)
  })
  normalize_edge_df(do.call(rbind, out))
}

collect_bridge_edges_for_children <- function(SP, sim_output, chr_info, child_iids) {
  pos_list <- attr(sim_output[[1L]], "tsForwardPosMeta", exact = TRUE)$posList
  bridge_list <- bridgeCollectSegGenFromSimOutput(
    SP,
    sim_output,
    chr_info = chr_info,
    pos_list = pos_list
  )
  if (length(bridge_list) == 0L) {
    return(data.frame(chr = integer(), child = integer(), parent = integer(),
                      left = numeric(), right = numeric(), stringsAsFactors = FALSE))
  }
  bridge_df <- do.call(rbind, bridge_list)
  bridge_edges <- data.frame(
    chr = as.integer(bridge_df$chr),
    child = as.integer(bridge_df$childId),
    parent = as.integer(bridge_df$parentId),
    left = as.numeric(bridge_df$left),
    right = as.numeric(bridge_df$right),
    stringsAsFactors = FALSE
  )
  bridge_edges <- bridge_edges[bridge_edges$child %in% child_iids, , drop = FALSE]
  normalize_edge_df(bridge_edges)
}

compare_direct_vs_bridge_edges <- function(recorder, SP, sim_output, chr_info) {
  child_iids <- unlist(lapply(sim_output[-1L], function(pop) as.integer(pop@iid)), use.names = FALSE)
  direct_edges <- collect_forward_edges_all_chr(recorder, child_iids = child_iids)
  bridge_edges <- collect_bridge_edges_for_children(
    SP = SP,
    sim_output = sim_output,
    chr_info = chr_info,
    child_iids = child_iids
  )

  checks <- data.frame(
    metric = c("n_edges", "chr_child_parent", "left", "right"),
    equal = c(
      nrow(direct_edges) == nrow(bridge_edges),
      identical(direct_edges[, c("chr", "child", "parent"), drop = FALSE],
                bridge_edges[, c("chr", "child", "parent"), drop = FALSE]),
      isTRUE(all.equal(direct_edges$left, bridge_edges$left, tolerance = 1e-6)),
      isTRUE(all.equal(direct_edges$right, bridge_edges$right, tolerance = 1e-6))
    ),
    stringsAsFactors = FALSE
  )

  list(
    checks = checks,
    direct_edges = direct_edges,
    bridge_edges = bridge_edges
  )
}

edge_comparison_summary <- function(cmp) {
  direct <- cmp$direct_edges
  bridge <- cmp$bridge_edges
  n <- min(nrow(direct), nrow(bridge))
  same_ids <- if (n == 0L) {
    nrow(direct) == nrow(bridge)
  } else {
    identical(direct[seq_len(n), c("chr", "child", "parent"), drop = FALSE],
              bridge[seq_len(n), c("chr", "child", "parent"), drop = FALSE])
  }
  max_left_diff <- if (n == 0L) NA_real_ else max(abs(direct$left[seq_len(n)] - bridge$left[seq_len(n)]))
  max_right_diff <- if (n == 0L) NA_real_ else max(abs(direct$right[seq_len(n)] - bridge$right[seq_len(n)]))

  data.frame(
    item = c(
      "direct forward edge rows",
      "bridge forward edge rows",
      "same chr/child/parent ordering",
      "max absolute left-coordinate difference",
      "max absolute right-coordinate difference"
    ),
    value = c(
      as.character(nrow(direct)),
      as.character(nrow(bridge)),
      as.character(same_ids),
      as.character(signif(max_left_diff, 6)),
      as.character(signif(max_right_diff, 6))
    ),
    stringsAsFactors = FALSE
  )
}

edge_example_table <- function(cmp, n = 8L) {
  direct <- cmp$direct_edges
  bridge <- cmp$bridge_edges
  n <- min(n, nrow(direct), nrow(bridge))
  if (n == 0L) {
    return(data.frame())
  }
  data.frame(
    row = seq_len(n),
    chr = direct$chr[seq_len(n)],
    child = direct$child[seq_len(n)],
    parent = direct$parent[seq_len(n)],
    direct_left = direct$left[seq_len(n)],
    bridge_left = bridge$left[seq_len(n)],
    left_diff = direct$left[seq_len(n)] - bridge$left[seq_len(n)],
    direct_right = direct$right[seq_len(n)],
    bridge_right = bridge$right[seq_len(n)],
    right_diff = direct$right[seq_len(n)] - bridge$right[seq_len(n)],
    stringsAsFactors = FALSE
  )
}

edge_example_table_by_chr <- function(cmp, n = 5L) {
  direct <- cmp$direct_edges
  out <- lapply(split(seq_len(nrow(direct)), direct$chr), function(idx) {
    cmp_chr <- list(
      direct_edges = cmp$direct_edges[idx, , drop = FALSE],
      bridge_edges = cmp$bridge_edges[idx, , drop = FALSE]
    )
    edge_example_table(cmp_chr, n = n)
  })
  do.call(rbind, out)
}


find_extra_recHistGen_example <- function(SP) {
  RHG <- SP$recHistGen
  RH <- SP$recHist
  if (is.null(RHG) || is.null(RH)) {
    return(NULL)
  }

  as_rec_matrix <- function(x) {
    if (is.null(x) || length(x) == 0L) {
      return(NULL)
    }
    x <- as.matrix(x)
    if (ncol(x) < 2L || nrow(x) == 0L) {
      return(NULL)
    }
    x
  }

  for (iid in seq_along(RHG)) {
    for (chr in seq_along(RHG[[iid]])) {
      gm <- as.numeric(SP$genMap[[chr]])
      for (hap in seq_along(RHG[[iid]][[chr]])) {
        rhg <- as_rec_matrix(RHG[[iid]][[chr]][[hap]])
        rh <- as_rec_matrix(RH[[iid]][[chr]][[hap]])
        if (is.null(rhg) || is.null(rh) || nrow(rhg) <= nrow(rh)) {
          next
        }

        next_marker <- findInterval(as.numeric(rhg[, 2]), gm) + 1L
        next_marker <- pmin(pmax(next_marker, 1L), length(gm))
        next_marker[seq_along(next_marker) == 1L] <- 1L
        duplicated_marker <- unique(next_marker[-1L][duplicated(next_marker[-1L])])
        if (length(duplicated_marker) == 0L) {
          next
        }

        marker <- duplicated_marker[[1L]]
        keep <- which(next_marker == marker)
        keep <- seq(max(1L, min(keep) - 1L), min(nrow(rhg), max(keep) + 1L))
        table_next_marker <- next_marker[keep]
        table_left_marker <- ifelse(
          table_next_marker > 1L,
          gm[table_next_marker - 1L],
          NA_real_
        )
        rhg_table <- data.frame(
          row = keep,
          source_hap = as.integer(rhg[keep, 1]),
          breakpoint_morgan = as.numeric(rhg[keep, 2]),
          next_marker = table_next_marker,
          left_marker_morgan = table_left_marker,
          right_marker_morgan = gm[table_next_marker],
          stringsAsFactors = FALSE
        )
        rh_table <- data.frame(
          row = seq_len(nrow(rh)),
          source_hap = as.integer(rh[, 1]),
          start_marker = as.integer(rh[, 2]),
          start_marker_morgan = gm[pmin(pmax(as.integer(rh[, 2]), 1L), length(gm))],
          stringsAsFactors = FALSE
        )

        return(list(
          iid = iid,
          chr = chr,
          hap = hap,
          marker = marker,
          rhg_table = rhg_table,
          rh_table = rh_table
        ))
      }
    }
  }
  NULL
}
```

## 3) Entry Point A: Start With `runMacsTS`

`runMacsTS()` returns a `MapPop` plus TS table pointers when `returnTs=TRUE`.
For the new direct recorder path, use `mac_out$pop` directly. The extra
file-backed `chr_info` below is only for the old bridge comparison functions,
because those functions still expect `chr_info[[cc]]$ts_path`.

```{r runMacsTS-founders}
set.seed(42)

# Step 1A: create founder tree sequences with runMacsTS.
mac_out <- AlphaSimR:::runMacsTS(
  nInd = 4,
  nChr = 2,
  segSites = c(60L, 120L),
  inbred = FALSE,
  ploidy = 2L,
  species = "GENERIC",
  mutationMode = "postTs",
  usePhysicalPositions = TRUE,
  nThreads = 1L,
  returnTs = TRUE
)

# Direct forward-recorder input: use the MapPop returned by runMacsTS.
mac_founder <- mac_out$pop

# Old bridge input: bridgeCollectSegGenFromSimOutput()/bridgeWriteTrees() still
# use chr_info[[cc]]$ts_path, so this validation-only helper writes the in-memory
# runMacsTS founder tables to files and reconstructs the bridge rate map.
mac_bridge_chr_info <- AlphaSimR:::runMacsTSBridgeChrInfo(
  mac_out,
  out_dir = out_dir,
  out_basename = "runMacsTS_founder"
)
```

### Forward Simulation With Direct TS Recorder

```{r runMacsTS-forward}
# Step 2A: run the forward simulator and let the direct TS recorder update
# tables during meiosis. Keep recHistGen only for validation against the old
# bridge path.
old_opts_mac <- options(
  AlphaSimR.tsForwardKeepSeg = FALSE,
  AlphaSimR.tsForwardKeepRecHistGen = TRUE
)

set.seed(42)
mac_SP <- SimParam$new(mac_founder)
mac_SP$nThreads <- 1L
mac_SP$quadProb <- 0
mac_SP$setSexes("yes_sys")
mac_SP$addTraitA(nQtlPerChr = 5, mean = 500, var = 450)
mac_SP$setTrackPed(TRUE)
mac_SP$setTrackRecGen(TRUE)

mac_basePop <- newPop(mac_founder, simParam = mac_SP)
mac_SP$setTrackTs(TRUE, founderPop = mac_basePop)
mac_basePop <- setPheno(mac_basePop, h2 = 0.5, simParam = mac_SP)

mac_sim_output <- list(mac_basePop)
for (cycle in seq_len(2L)) {
  prev <- mac_sim_output[[cycle]]
  chosen <- selectInd(pop = prev, nInd = min(6L, nInd(prev)), use = "gv", simParam = mac_SP)
  offspring <- randCross(pop = chosen, nCrosses = 2L, nProgeny = 5L, simParam = mac_SP)
  offspring <- setPheno(pop = offspring, h2 = 0.5, simParam = mac_SP)
  mac_sim_output[[cycle + 1L]] <- offspring
}

# Only for debugging and validation; for users, run tsForwardFinalizeFromSimParam() to write .trees files (see below).
mac_recorder <- attr(mac_SP, "tsForwardRecorder", exact = TRUE)
mac_recorder
stopifnot(inherits(mac_recorder, "tsForwardRecorder"))
# restore options
options(old_opts_mac)
```

### Compare Direct Recorder With The Old Bridge Path

Each row in these edge tables means: for one child haplotype, the interval
`[left, right)` on chromosome `chr` inherited from `parent`. The direct recorder
appends these intervals directly to the TS edge table during crossing. The old
bridge path reconstructs the same intervals later from `recHistGen`.


The comparison is about the forward-added inheritance edges for offspring
created after the founder generation. A single edge row means:

- `chr`: chromosome index in AlphaSimR,
- `child`: AlphaSimR individual id for the offspring,
- `parent`: AlphaSimR individual id for the parent that contributed this interval,
- `left`, `right`: the half-open tree-sequence interval `[left, right)`.

The direct recorder writes these intervals to the TS edge table during crossing.
The bridge path reconstructs them later from `recHistGen` using
`bridgeCollectSegGenFromSimOutput()`. The summary table reports both edge-row
counts and the maximum absolute coordinate difference. The example table then
prints paired direct/bridge rows so the agreement can be inspected directly,
not just as a `TRUE`/`FALSE` result.

Expected result:

- direct and bridge edge counts should be the same,
- `same chr/child/parent ordering` should be `TRUE`,
- max absolute left/right differences should be zero or within floating-point
  tolerance,
- the direct and bridge `.trees` files should have matching forward-added edge
  structure and can be inspected further with Python/tskit.

```{r runMacsTS-compare}
mac_cmp <- compare_direct_vs_bridge_edges(
  recorder = mac_recorder,
  SP = mac_SP,
  sim_output = mac_sim_output,
  chr_info = mac_bridge_chr_info
)

knitr::kable(edge_comparison_summary(mac_cmp))
#knitr::kable(edge_example_table(mac_cmp, n = 30L))
knitr::kable(edge_example_table_by_chr(mac_cmp, n = 5L), digits = 3,
  format.args = list(scientific = FALSE, nsmall = 3)
)
```

### Write Comparable Tree Files

```{r runMacsTS-write-trees}
mac_direct_paths <- tsForwardFinalizeFromSimParam(
  mac_SP,
  out_dir = out_dir,
  out_basename = "runMacsTS_direct",
  clear = FALSE
)

# keep founder nodes; nodes in the current generation and nodes in their pedigree
mac_direct_simplified_paths <- tsForwardFinalizeFromSimParam(
  mac_SP,
  out_dir = out_dir,
  out_basename = "runMacsTS_direct_simplified_keep_unary",
  simplify = TRUE,
  keep_unary = TRUE,
  clear = FALSE
)

mac_bridge_list <- bridgeCollectSegGenFromSimOutput(
  mac_SP,
  mac_sim_output,
  chr_info = mac_bridge_chr_info,
  pos_list = attr(mac_sim_output[[1L]], "tsForwardPosMeta", exact = TRUE)$posList
)
mac_bridge_df <- do.call(rbind, mac_bridge_list)
bridgeWriteTrees(
  chr_info = mac_bridge_chr_info,
  edgeDf = mac_bridge_df,
  SP = mac_SP,
  out_dir = out_dir,
  out_basename = "runMacsTS_bridge"
)
mac_bridge_paths <- file.path(out_dir, paste0("runMacsTS_bridge_chr", seq_along(mac_bridge_chr_info) - 1L, ".trees"))

mac_tree_paths <- data.frame(
  chr = seq_along(mac_bridge_chr_info),
  direct_path = mac_direct_paths,
  direct_simplified_path = mac_direct_simplified_paths,
  bridge_path = mac_bridge_paths,
  direct_exists = file.exists(mac_direct_paths),
  direct_simplified_exists = file.exists(mac_direct_simplified_paths),
  bridge_exists = file.exists(mac_bridge_paths),
  stringsAsFactors = FALSE
)
mac_tree_paths
```

## 4) Entry Point B: Start With Existing msprime Tree Files

This mirrors the setup in `dev/notesRealBreakpoints.md`. The recombination map
is supplied explicitly in `chr_info`. Here the rates are deliberately higher
than the original msprime examples, on the order of `1e-6`, so the forward
simulation creates more crossing-over events, including events between retained
markers.

```{r msprime-founders}
L1 <- 1e6
L2 <- 2e6
msprime_chr_info <- list(
  list(
    ts_path = paste0(out_dir, "/msprime_chr0.trees"),
    breaks = c(0, L1 / 2, L1),
    rates = c(5e-5, 5e-6),
    segSites = 60L
  ),
  list(
    ts_path = paste0(out_dir, "/msprime_chr1.trees"),
    breaks = c(0, L2 / 3, 2 * L2 / 3, L2),
    rates = c(5e-7, 5e-6, 5e-7),
    segSites = 155L
  )
)

stopifnot(all(file.exists(vapply(msprime_chr_info, `[[`, character(1), "ts_path"))))

msprime_founder <- asMapPop(
  chr_info = msprime_chr_info,
  inbred = FALSE,
  ploidy = 2L
)
```

### Forward Simulation With Direct TS Recorder

```{r msprime-forward}
# Step 2B: run the same forward TS recording process, now starting from
# existing msprime tree files.
old_opts_msprime <- options(
  AlphaSimR.tsForwardKeepSeg = FALSE,
  AlphaSimR.tsForwardKeepRecHistGen = TRUE
)

set.seed(42)
msprime_SP <- SimParam$new(msprime_founder)
msprime_SP$nThreads <- 1L
msprime_SP$quadProb <- 0
msprime_SP$setSexes("no")
msprime_SP$addTraitA(nQtlPerChr = 5, mean = 500, var = 450)
msprime_SP$setTrackPed(TRUE)
msprime_SP$setTrackRecGen(TRUE)

msprime_basePop <- newPop(msprime_founder, simParam = msprime_SP)
msprime_SP$setTrackTs(TRUE, founderPop = msprime_basePop)
msprime_basePop <- setPheno(msprime_basePop, h2 = 0.5, simParam = msprime_SP)

msprime_sim_output <- list(msprime_basePop)
for (cycle in seq_len(2L)) {
  prev <- msprime_sim_output[[cycle]]
  chosen <- selectInd(pop = prev, nInd = min(2L, nInd(prev)), use = "gv", simParam = msprime_SP)
  offspring <- randCross(pop = chosen, nCrosses = 1L, nProgeny = 5L, simParam = msprime_SP)
  offspring <- setPheno(pop = offspring, h2 = 0.5, simParam = msprime_SP)
  msprime_sim_output[[cycle + 1L]] <- offspring
}

msprime_recorder <- attr(msprime_SP, "tsForwardRecorder", exact = TRUE)
stopifnot(inherits(msprime_recorder, "tsForwardRecorder"))
options(old_opts_msprime)
```

### Compare Direct Recorder With The Old Bridge Path

```{r msprime-compare}
msprime_cmp <- compare_direct_vs_bridge_edges(
  recorder = msprime_recorder,
  SP = msprime_SP,
  sim_output = msprime_sim_output,
  chr_info = msprime_chr_info
)

knitr::kable(edge_comparison_summary(msprime_cmp))
#knitr::kable(edge_example_table(msprime_cmp, n = 10L))
knitr::kable(edge_example_table_by_chr(msprime_cmp, n = 5L), digits = 3,
  format.args = list(scientific = FALSE, nsmall = 3)
)
```

### Recombination Breakpoints Between Markers

The comparison above uses `bridgeCollectSegGenFromSimOutput()`, so the bridge
side is reconstructed from `recHistGen`. When the direct-recorder edges match
that bridge output, it means the forward recorder is using the same breakpoint
resolution as `recHistGen`.

The marker-index `recHist` representation can still miss double crossing-over
events that happen between the same two retained markers, because the marker
ancestry can return to the same state before the next marker.

```{r msprime-double-crossover}
double_example <- find_extra_recHistGen_example(msprime_SP)
if (is.null(double_example)) {
  cat("No between-marker double-crossover example found in this seeded run.\n")
} else {
  cat(
    "Check ind", double_example$iid,
    "; chr", double_example$chr,
    "; hap", double_example$hap, "\n"
  )

  RHG <- msprime_SP$recHistGen
  RH <- msprime_SP$recHist
  gm <- msprime_SP$genMap[[double_example$chr]]

  rh <- as.matrix(RH[[double_example$iid]][[double_example$chr]][[double_example$hap]])
  rhg <- as.matrix(RHG[[double_example$iid]][[double_example$chr]][[double_example$hap]])

  cat(
    "recHistGen has", nrow(rhg) - nrow(rh),
    "more recorded recombination events than recHist.\n"
  )

  cat("\nrecHist: marker-index ancestry changes\n")
  print(rh)
  cat("\nrecHistGen: actual breakpoint positions in Morgan\n")
  print(rhg)

  x <- as.numeric(rhg[, 2])
  left <- findInterval(x, gm)
  left <- pmax(left, 1L)
  right <- pmin(left + 1L, length(gm))

  recHistGen_intervals <- data.frame(
    x = x,
    left_i = left,
    left_v = gm[left],
    right_i = right,
    right_v = gm[right],
    stringsAsFactors = FALSE
  )

  cat("\nSNP interval containing each recHistGen breakpoint\n")
  print(recHistGen_intervals)
}
```

Rows in the interval table with the same `left_i` and `right_i` are multiple
breakpoints between the same two retained SNPs. These are retained by
`recHistGen`, and therefore by the forward TS recorder, but they are collapsed
in marker-index `recHist`.

### Write Comparable Tree Files

```{r msprime-write-trees}
msprime_direct_paths <- tsForwardFinalizeFromSimParam(
  msprime_SP,
  out_dir = out_dir,
  out_basename = "msprime_direct",
  clear = FALSE
)

# keep founder nodes; nodes in the current generation and nodes in their pedigree
msprime_direct_simplified_paths <- tsForwardFinalizeFromSimParam(
  msprime_SP,
  out_dir = out_dir,
  out_basename = "msprime_direct_simplified_keep_unary",
  simplify = TRUE,
  keep_unary = TRUE,
  clear = FALSE
)

msprime_bridge_list <- bridgeCollectSegGenFromSimOutput(
  msprime_SP,
  msprime_sim_output,
  chr_info = msprime_chr_info,
  pos_list = attr(msprime_sim_output[[1L]], "tsForwardPosMeta", exact = TRUE)$posList
)
msprime_bridge_df <- do.call(rbind, msprime_bridge_list)
bridgeWriteTrees(
  chr_info = msprime_chr_info,
  edgeDf = msprime_bridge_df,
  SP = msprime_SP,
  out_dir = out_dir,
  out_basename = "msprime_bridge"
)
msprime_bridge_paths <- file.path(out_dir, paste0("msprime_bridge_chr", seq_along(msprime_chr_info) - 1L, ".trees"))

msprime_tree_paths <- data.frame(
  chr = seq_along(msprime_chr_info),
  direct_path = msprime_direct_paths,
  direct_simplified_path = msprime_direct_simplified_paths,
  bridge_path = msprime_bridge_paths,
  direct_exists = file.exists(msprime_direct_paths),
  direct_simplified_exists = file.exists(msprime_direct_simplified_paths),
  bridge_exists = file.exists(msprime_bridge_paths),
  stringsAsFactors = FALSE
)

```

```{r combine-tree-paths}
tree_path_cols <- c(
  "chr",
  "direct_path",
  "direct_simplified_path",
  "bridge_path",
  "direct_exists",
  "direct_simplified_exists",
  "bridge_exists"
)
align_tree_path_cols <- function(x) {
  missing_cols <- setdiff(tree_path_cols, names(x))
  for (col in missing_cols) {
    x[[col]] <- NA
  }
  x[, tree_path_cols, drop = FALSE]
}
all_tree_paths <- rbind(
  align_tree_path_cols(mac_tree_paths),
  align_tree_path_cols(msprime_tree_paths)
)
all_tree_paths
```

## 5) Files To Compare In Python And Interpret

Use the following Python code outside R, or run the chunk after configuring a
Python environment with `tskit` installed.
```{r py-setup}
# Use your own path here!
reticulate::use_python("/usr/bin/python3", required = TRUE)
library(reticulate)
reticulate::py_config()
```

```{r python-paths}
python_pairs <- data.frame(
  label = c("runMacsTS chr0", "runMacsTS chr1", "msprime chr0", "msprime chr1"),
  direct_path = normalizePath(all_tree_paths$direct_path, mustWork = FALSE),
  bridge_path = normalizePath(all_tree_paths$bridge_path, mustWork = FALSE),
  stringsAsFactors = FALSE
)
library(reticulate)
pairs_obj <- lapply(seq_len(nrow(python_pairs)), function(i) {
  as.list(python_pairs[i, , drop = FALSE])
})
reticulate::py_run_string("pairs = r.pairs_obj")
```

```{python compare-tree-files}
from pathlib import Path
import tskit

required_fields = {"direct_path", "bridge_path"}
missing = required_fields - set(pairs[0])
if missing:
    raise ValueError(f"python_pairs is missing {sorted(missing)}. Rerun the python-paths R chunk.")

def direct_simplified_path(direct_path):
    path = Path(direct_path)
    return path.with_name(
        path.name.replace("_direct_chr", "_direct_simplified_keep_unary_chr")
    )

for row in pairs:
    print(row)
    direct = tskit.load(row["direct_path"])
    direct_simplified = tskit.load(direct_simplified_path(row["direct_path"]))
    bridge = tskit.load(row["bridge_path"])
    print(row["label"])
    print("  order: direct, bridge, direct_simplified_keep_unary")
    print("  nodes:", direct.num_nodes, bridge.num_nodes, direct_simplified.num_nodes)
    print("  edges:", direct.num_edges, bridge.num_edges, direct_simplified.num_edges)
    print("  trees:", direct.num_trees, bridge.num_trees, direct_simplified.num_trees)
    print("  sites:", direct.num_sites, bridge.num_sites, direct_simplified.num_sites)
```

Print one tree from each differently generated result. The simplified direct
files are generated above by
`tsForwardFinalizeFromSimParam(..., simplify = TRUE, keep_unary = TRUE)`.

```{python print-example-trees}

for row in pairs:
    if row["label"] not in {"msprime chr0"}:
        continue
    for method in ["direct", "bridge"]:
        ts_path = Path(row[f"{method}_path"])
        ts = tskit.load(ts_path)

        print(f"\n{row['label']} - {method}")
        print("original counts")
        print("  trees:", ts.num_trees)
        print("  nodes:", ts.num_nodes)
        print("  edges:", ts.num_edges)
        print(ts.first().draw_text())

    direct = tskit.load(row["direct_path"])
    simplified_path = direct_simplified_path(row["direct_path"])
    simplified = tskit.load(simplified_path)
    print(f"\n{row['label']} - direct simplified by AlphaSimR finalize, keep_unary=True")
    print("  path:", simplified_path)
    print("  trees:", simplified.num_trees, "reduced by", direct.num_trees - simplified.num_trees)
    print("  nodes:", simplified.num_nodes, "reduced by", direct.num_nodes - simplified.num_nodes)
    print("  edges:", simplified.num_edges, "reduced by", direct.num_edges - simplified.num_edges)
    print(simplified.first().draw_text())

# see the information of the only offspring of founder node 1    
print(direct.node(270))
print(simplified.node(24))

print(direct.individual(17))
print(simplified.individual(14))
```

Print and draw pedigrees from the same files:

```{python print-pedigrees}
from pathlib import Path
import tskit
from matplotlib import pyplot as plt
import networkx as nx

def draw_pedigree(ped_ts, title):
    plt.figure(figsize=(8, 5))

    G = nx.DiGraph()
    for ind in ped_ts.individuals():
        if len(ind.nodes) == 0:
            continue
        time = ped_ts.node(ind.nodes[0]).time
        pop = ped_ts.node(ind.nodes[0]).population
        G.add_node(ind.id, time=time, population=pop)
        for p in ind.parents:
            if p != tskit.NULL:
                G.add_edge(ind.id, p)

    pos = nx.multipartite_layout(G, subset_key="time", align="horizontal")
    colours = plt.rcParams["axes.prop_cycle"].by_key()["color"]
    node_colours = [
        colours[attr["population"] % len(colours)]
        for attr in G.nodes.values()
    ]

    nx.draw_networkx(G, pos, with_labels=True, node_color=node_colours)
    plt.title(title)
    plt.show()
    plt.close()

print(pairs)
for row in pairs:
    if row["label"] not in ["msprime chr1", "runMacsTS chr1"]:
        continue
    #print(row)
    paths = {
        "direct": row["direct_path"],
        "bridge": row["bridge_path"],
        "direct_simplified": direct_simplified_path(row["direct_path"])
    }

    for method, path in paths.items():
        print(row)
        ts = tskit.load(path)
        print(f"\n{row['label']} - {method}")
        draw_pedigree(ts, f"{row['label']} - {method}")
```

## 6) Test Recombination Map Handling

This section checks how recombination-map information is carried from founder
generation into the forward simulation and the direct TS recorder. When
`runMacsTS()` is run with a MaCS `-R` hotspot file, the resulting founder keeps
that map metadata, so the forward recorder can use it without additional user
input. When `asMapPop(chr_info = list(..., breaks = ..., rates = ...))` is used,
the explicit `breaks/rates` input controls the founder `genMap` used by the
forward simulation.

We test this because there are two separate sources of recombination-map
information: the map used to generate the MaCS ancestry, and the map stored on
the founder for later AlphaSimR forward recombination. These examples check that
the forward edge-start distribution follows the map available to the forward
simulation, including the case where constant-rate MaCS ancestry is rebuilt with
an explicit non-constant map for the forward stage.

The four cases are:

1. constant recombination map from the normal `runMacsTS()` path,
2. constant-rate `runMacsTS()` ancestry rebuilt with explicit hotspot
   `breaks/rates` for the forward simulation,
3. MaCS `-R` hotspot file parsed automatically by `runMacsTS()`,
4. MaCS `-R` ancestry tables rebuilt with explicit `breaks/rates` through
   `asMapPop(chr_info = list(tables = ...))`.

The dashed lines in the plot show the expected number of edge starts per
window: number of transmitted haplotypes times the window's genetic length.

```{r runmacts-recombination-map-edge-windows}
recomb_test_bp <- 1e8
recomb_test_Ne <- 100
recomb_test_mut_rate <- 2.5e-8
recomb_test_rec_rate <- 1e-8
recomb_test_seg_sites <- 80L
recomb_test_window_bp <- 1e7
recomb_test_n_forward_gen <- 2L
recomb_test_n_crosses <- 3L
recomb_test_n_progeny <- 5L
recomb_test_ploidy <- 2L
recomb_test_hotspot_lines <- c("0.2 0.3 5", "0.6 0.8 0.1")
recomb_test_hotspot_path <- file.path(out_dir, "runMacsTS_hotspots_for_window_counts.txt")

edge_start_window_counts <- function(edges, seq_len, window_bp = 1e7) {
  starts <- seq(0, seq_len - window_bp, by = window_bp)
  ends <- pmin(starts + window_bp, seq_len)
  edge_starts <- edges$left[edges$left > 0 & edges$left < seq_len]
  data.frame(
    window = seq_along(starts),
    window_start = starts,
    window_end = ends,
    n_edge_starts = vapply(seq_along(starts), function(i) {
      sum(edge_starts >= starts[[i]] & edge_starts < ends[[i]])
    }, integer(1)),
    stringsAsFactors = FALSE
  )
}

run_forward_edge_windows <- function(founder, rep_id, case, seq_len, window_bp = 1e7) {
  SP <- SimParam$new(founder)
  SP$nThreads <- 1L
  SP$quadProb <- 0
  SP$setTrackPed(TRUE)

  base_pop <- newPop(founder, simParam = SP)
  SP$setTrackTs(TRUE, founderPop = base_pop)

  set.seed(10000L + rep_id)
  sim_output <- list(base_pop)
  for (gen in seq_len(recomb_test_n_forward_gen)) {
    sim_output[[gen + 1L]] <- randCross(
      pop = sim_output[[gen]],
      nCrosses = recomb_test_n_crosses,
      nProgeny = recomb_test_n_progeny,
      simParam = SP
    )
  }

  recorder <- attr(SP, "tsForwardRecorder", exact = TRUE)
  child_iids <- unlist(lapply(sim_output[-1L], function(pop) as.integer(pop@iid)), use.names = FALSE)
  edges <- collect_forward_edges_all_chr(recorder, child_iids = child_iids)
  out <- edge_start_window_counts(edges, seq_len = seq_len, window_bp = window_bp)
  out$rep <- rep_id
  out$case <- case
  out
}

expected_edge_start_windows <- function(seq_len, window_bp, breaks, rates,
                                        n_meioses, expected_map) {
  starts <- seq(0, seq_len - window_bp, by = window_bp)
  ends <- pmin(starts + window_bp, seq_len)
  genetic_span <- rateMap2cumMorgan(ends, breaks, rates) -
    rateMap2cumMorgan(starts, breaks, rates)
  data.frame(
    window = seq_along(starts),
    window_start = starts,
    window_end = ends,
    window_mid = (starts + ends) / 2,
    expected_edge_starts = n_meioses * genetic_span,
    expected_map = expected_map,
    stringsAsFactors = FALSE
  )
}

runmacts_recombination_map_replicate <- function(rep_id) {
  bp <- recomb_test_bp
  Ne <- recomb_test_Ne
  mut_rate <- recomb_test_mut_rate
  rec_rate <- recomb_test_rec_rate
  seg_sites <- recomb_test_seg_sites

  dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  writeLines(recomb_test_hotspot_lines, recomb_test_hotspot_path)

  old_opts <- options(
    AlphaSimR.tsForwardAttachMetadata = TRUE,
    AlphaSimR.tsForwardKeepSeg = FALSE,
    AlphaSimR.tsForwardKeepRecHistGen = FALSE
  )
  on.exit(options(old_opts), add = TRUE)

  manual_command <- paste(
    bp,
    "-t", 4 * Ne * mut_rate,
    "-r", 4 * Ne * rec_rate,
    "-R", recomb_test_hotspot_path
  )

  explicit_forward_map <- AlphaSimR:::.runMacsTS_map_from_hotspots(
    path = recomb_test_hotspot_path,
    nChr = 1,
    seqLen = bp,
    genLen = rec_rate * bp,
    usePhysicalPositions = TRUE
  )

  constant_seed <- as.integer(20000L + rep_id)
  constant_mut_seed <- as.integer(25000L + rep_id)
  constant_site_seed <- as.integer(30000L + rep_id)

  run_constant_macs <- function() {
    AlphaSimR:::runMacsTS(
      nInd = 4,
      nChr = 1,
      segSites = seg_sites,
      inbred = FALSE,
      ploidy = recomb_test_ploidy,
      species = "GENERIC",
      mutationMode = "postTs",
      usePhysicalPositions = TRUE,
      nThreads = 1L,
      seed = constant_seed,
      mutSeed = constant_mut_seed,
      siteSamplingSeed = constant_site_seed,
      returnTs = TRUE
    )
  }

  # Group 1 source: constant-rate MaCS ancestry with the default constant
  # forward map retained from runMacsTS().
  constant_out <- run_constant_macs()

  # Group 2 source: an independent copy of the same constant-rate MaCS
  # ancestry, rebuilt below with explicit hotspot breaks/rates.
  constant_out_custom_map <- run_constant_macs()

  # Group 2: constant-rate MaCS ancestry, but explicit hotspot breaks/rates
  # are supplied to asMapPop() so the forward simulation uses the custom map.
  constant_founder_custom_forward_map <- asMapPop(
    chr_info = list(
      tables = constant_out_custom_map$tables,
      breaks = explicit_forward_map$breaks,
      rates = explicit_forward_map$rates
    ),
    inbred = FALSE,
    ploidy = recomb_test_ploidy,
    segSites = seg_sites,
    site_sampling_seed = constant_site_seed,
    nThreads = 1L
  )

  hotspot_seed <- as.integer(40000L + rep_id)
  hotspot_mut_seed <- as.integer(45000L + rep_id)
  hotspot_site_seed <- as.integer(50000L + rep_id)

  run_hotspot_macs <- function() {
    AlphaSimR:::runMacsTS(
      nInd = 4,
      nChr = 1,
      segSites = seg_sites,
      inbred = FALSE,
      ploidy = recomb_test_ploidy,
      manualCommand = manual_command,
      manualGenLen = rec_rate * bp,
      mutationMode = "postTs",
      usePhysicalPositions = TRUE,
      nThreads = 1L,
      seed = hotspot_seed,
      mutSeed = hotspot_mut_seed,
      siteSamplingSeed = hotspot_site_seed,
      returnTs = TRUE
    )
  }

  # Group 3 source: MaCS -R ancestry where runMacsTS() automatically retains
  # the hotspot map metadata for the forward simulation.
  hotspot_out_auto <- run_hotspot_macs()

  # Group 4 source: an independent copy of the same MaCS -R ancestry, rebuilt
  # below with explicit hotspot breaks/rates.
  hotspot_out_explicit <- run_hotspot_macs()

  hotspot_pos_meta <- attr(hotspot_out_explicit$pop, "tsForwardPosMeta", exact = TRUE)

  # Group 4: MaCS -R ancestry rebuilt with the same explicit hotspot
  # breaks/rates through asMapPop().
  explicit_founder <- asMapPop(
    chr_info = list(
      tables = hotspot_out_explicit$tables,
      breaks = hotspot_pos_meta$breaksList,
      rates = hotspot_pos_meta$ratesList
    ),
    inbred = FALSE,
    ploidy = recomb_test_ploidy,
    segSites = seg_sites,
    site_sampling_seed = as.integer(50000L + rep_id),
    nThreads = 1L
  )

  do.call(rbind, list(
    # Group 1: constant-rate MaCS ancestry and constant-rate forward map.
    run_forward_edge_windows(
      founder = constant_out$pop,
      rep_id = rep_id,
      case = "constant",
      seq_len = bp
    ),
    # Group 2: constant-rate MaCS ancestry and explicit hotspot forward map.
    run_forward_edge_windows(
      founder = constant_founder_custom_forward_map,
      rep_id = rep_id,
      case = "constant_ancestry_explicit_forward_map",
      seq_len = bp
    ),
    # Group 3: MaCS -R ancestry and automatically retained hotspot map.
    run_forward_edge_windows(
      founder = hotspot_out_auto$pop,
      rep_id = rep_id,
      case = "macs_R_auto",
      seq_len = bp
    ),
    # Group 4: MaCS -R ancestry and explicit hotspot forward map.
    run_forward_edge_windows(
      founder = explicit_founder,
      rep_id = rep_id,
      case = "macs_R_explicit_breaks_rates",
      seq_len = bp
    )
  ))
}

edge_window_reps <- do.call(rbind, lapply(seq_len(50L), runmacts_recombination_map_replicate))

expected_n_meioses <- recomb_test_n_forward_gen *
  recomb_test_n_crosses *
  recomb_test_n_progeny *
  recomb_test_ploidy

expected_hotspot_map <- AlphaSimR:::.runMacsTS_map_from_hotspots(
  path = recomb_test_hotspot_path,
  nChr = 1,
  seqLen = recomb_test_bp,
  genLen = recomb_test_rec_rate * recomb_test_bp,
  usePhysicalPositions = TRUE
)

expected_edge_windows <- rbind(
  expected_edge_start_windows(
    seq_len = recomb_test_bp,
    window_bp = recomb_test_window_bp,
    breaks = c(0, recomb_test_bp),
    rates = c(recomb_test_rec_rate),
    n_meioses = expected_n_meioses,
    expected_map = "expected_constant_rate"
  ),
  expected_edge_start_windows(
    seq_len = recomb_test_bp,
    window_bp = recomb_test_window_bp,
    breaks = expected_hotspot_map$breaks[[1]],
    rates = expected_hotspot_map$rates[[1]],
    n_meioses = expected_n_meioses,
    expected_map = "expected_hotspot_map"
  )
)

edge_window_summary <- aggregate(
  n_edge_starts ~ case + window + window_start + window_end,
  data = edge_window_reps,
  FUN = function(x) c(mean = mean(x), sd = stats::sd(x))
)
edge_window_summary <- do.call(data.frame, edge_window_summary)
names(edge_window_summary)[names(edge_window_summary) == "n_edge_starts.mean"] <- "mean_edge_starts"
names(edge_window_summary)[names(edge_window_summary) == "n_edge_starts.sd"] <- "sd_edge_starts"
edge_window_summary <- edge_window_summary[
  order(edge_window_summary$case, edge_window_summary$window),
  ,
  drop = FALSE
]
rownames(edge_window_summary) <- NULL

edge_window_summary
```

```{r runmacts-recombination-map-edge-start-plot}
edge_window_summary$window_mid <- (
  edge_window_summary$window_start + edge_window_summary$window_end
) / 2

edge_cases <- unique(edge_window_summary$case)
edge_cols <- setNames(c("#1B9E77", "#D95F02", "#7570B3", "#E7298A")[seq_along(edge_cases)], edge_cases)
expected_cols <- c(
  expected_constant_rate = "#555555",
  expected_hotspot_map = "#000000"
)
window_width <- min(diff(sort(unique(edge_window_summary$window_mid))))
edge_offsets <- setNames(
  seq(-0.18, 0.18, length.out = length(edge_cases)) * window_width,
  edge_cases
)

min_positive_y <- min(
  c(
    edge_window_summary$mean_edge_starts[edge_window_summary$mean_edge_starts > 0],
    expected_edge_windows$expected_edge_starts[expected_edge_windows$expected_edge_starts > 0]
  ),
  na.rm = TRUE
)
if (!is.finite(min_positive_y)) {
  min_positive_y <- 1
}
edge_window_summary$plot_x <- edge_window_summary$window_mid +
  edge_offsets[as.character(edge_window_summary$case)]
edge_window_summary$ymin <- pmax(
  edge_window_summary$mean_edge_starts - edge_window_summary$sd_edge_starts,
  min_positive_y / 2
)
edge_window_summary$ymax <- pmax(
  edge_window_summary$mean_edge_starts + edge_window_summary$sd_edge_starts,
  min_positive_y / 2
)

ggplot(
  edge_window_summary,
  aes(
    x = plot_x,
    y = mean_edge_starts,
    color = case,
    group = case
  )
) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_line(
    data = expected_edge_windows,
    aes(
      x = window_mid,
      y = expected_edge_starts,
      color = expected_map,
      group = expected_map
    ),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 0.9
  ) +
  geom_point(
    data = expected_edge_windows,
    aes(
      x = window_mid,
      y = expected_edge_starts,
      color = expected_map
    ),
    inherit.aes = FALSE,
    shape = 95,
    size = 5
  ) +
  scale_color_manual(values = c(edge_cols, expected_cols)) +
  scale_y_log10(
    limits = c(0.1, NA),
    breaks = c(0.6, 6, 30),
    labels = c("0.6", "6", "30")
  ) +
  labs(
    x = "Genome position (bp)",
    y = "Mean number of new edge starts, log10 scale",
    color = "Case"
  ) +
  theme_bw()
```

## 7) Standard User Workflow

This is the clean forward-recorder workflow without bridge validation,
`recHistGen`, saved `tsSegGen`, or comparison helpers. It shows both standard
entry points: start from `runMacsTS()` or start from existing msprime `.trees`
files. In both cases, the user-facing process is:

1. create founders,
2. initialize `SimParam` and the forward TS recorder,
3. run forward simulation,
4. write final `.trees` files.

```{r standard-forward-recorder-runmacts}
out_dir <- 'testData'

standard_runmacts_tree_paths <- local({
  old_opts_standard <- options(
    AlphaSimR.tsForwardAttachMetadata = TRUE,
    AlphaSimR.tsForwardKeepSeg = FALSE,
    AlphaSimR.tsForwardKeepRecHistGen = FALSE
  )
  on.exit(options(old_opts_standard), add = TRUE)

  set.seed(42)

  # 1. Create founders from runMacsTS.
  standard_mac_out <- AlphaSimR:::runMacsTS(
    nInd = 4,
    nChr = 2,
    segSites = c(60L, 120L),
    inbred = FALSE,
    ploidy = 2L,
    species = "GENERIC",
    mutationMode = "postTs",
    usePhysicalPositions = TRUE,
    nThreads = 1L,
    returnTs = TRUE
  )
  standard_founder <- standard_mac_out$pop

  # 2. Initialize AlphaSimR and the forward TS recorder.
  standard_SP <- SimParam$new(standard_founder)
  standard_SP$nThreads <- 1L
  standard_SP$quadProb <- 0
  standard_SP$setTrackPed(TRUE)

  standard_base_pop <- newPop(standard_founder, simParam = standard_SP)
  standard_SP$setTrackTs(TRUE, founderPop = standard_base_pop)

  # 3. Run the forward simulation. The recorder updates TS tables during meiosis.
  standard_pop <- standard_base_pop
  n_forward_gen <- 2L
  for (gen in seq_len(n_forward_gen)) {
    standard_pop <- randCross(
      pop = standard_pop,
      nCrosses = 2L,
      nProgeny = 5L,
      simParam = standard_SP
    )
  }

  # 4. Write final tree files.
  standard_out_dir <- file.path(out_dir)
  dir.create(standard_out_dir, recursive = TRUE, showWarnings = FALSE)
  standard_original_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
    standard_SP,
    out_dir = standard_out_dir,
    out_basename = "standard_runMacsTS_forward_recorder",
    clear = FALSE
  )
  standard_simplified_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
    standard_SP,
    out_dir = standard_out_dir,
    out_basename = "standard_runMacsTS_forward_recorder_simplified",
    simplify = TRUE,
    keep_unary = TRUE,
    clear = TRUE
  )
  list(original = standard_original_paths, simplified = standard_simplified_paths)
})

standard_runmacts_tree_paths
```

```{r standard-forward-recorder-msprime}
standard_msprime_tree_paths <- local({
  old_opts_standard <- options(
    AlphaSimR.tsForwardAttachMetadata = TRUE,
    AlphaSimR.tsForwardKeepSeg = FALSE,
    AlphaSimR.tsForwardKeepRecHistGen = FALSE
  )
  on.exit(options(old_opts_standard), add = TRUE)

  set.seed(42)

  # 1. Create founders from existing msprime tree files.
  # The recombination map must be supplied with external tree-sequence files.
  standard_L1 <- 1e6
  standard_L2 <- 2e6
  standard_msprime_chr_info <- list(
    list(
      ts_path = paste0(out_dir, "/msprime_chr0.trees"),
      breaks = c(0, standard_L1 / 2, standard_L1),
      rates = c(5e-5, 5e-6),
      segSites = 60L
    ),
    list(
      ts_path = paste0(out_dir, "/msprime_chr1.trees"),
      breaks = c(0, standard_L2 / 3, 2 * standard_L2 / 3, standard_L2),
      rates = c(5e-7, 5e-6, 5e-7),
      segSites = 155L
    )
  )
  stopifnot(all(file.exists(vapply(
    standard_msprime_chr_info, `[[`, character(1), "ts_path"
  ))))

  standard_msprime_founder <- asMapPop(
    chr_info = standard_msprime_chr_info,
    inbred = FALSE,
    ploidy = 2L
  )

  # 2. Initialize AlphaSimR and the forward TS recorder.
  standard_msprime_SP <- SimParam$new(standard_msprime_founder)
  standard_msprime_SP$nThreads <- 1L
  standard_msprime_SP$quadProb <- 0
  standard_msprime_SP$setTrackPed(TRUE)

  standard_msprime_base_pop <- newPop(
    standard_msprime_founder,
    simParam = standard_msprime_SP
  )
  standard_msprime_SP$setTrackTs(TRUE, founderPop = standard_msprime_base_pop)

  # 3. Run the forward simulation. The recorder updates TS tables during meiosis.
  standard_msprime_pop <- standard_msprime_base_pop
  n_forward_gen <- 2L
  for (gen in seq_len(n_forward_gen)) {
    standard_msprime_pop <- randCross(
      pop = standard_msprime_pop,
      nCrosses = 2L,
      nProgeny = 5L,
      simParam = standard_msprime_SP
    )
  }

  # 4. Write final tree files.
  standard_out_dir <- file.path(out_dir)
  dir.create(standard_out_dir, recursive = TRUE, showWarnings = FALSE)
  standard_msprime_original_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
    standard_msprime_SP,
    out_dir = standard_out_dir,
    out_basename = "standard_msprime_forward_recorder",
    clear = FALSE
  )
  standard_msprime_simplified_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
    standard_msprime_SP,
    out_dir = standard_out_dir,
    out_basename = "standard_msprime_forward_recorder_simplified",
    simplify = TRUE,
    keep_unary = TRUE,
    clear = TRUE
  )
  list(
    original = standard_msprime_original_paths,
    simplified = standard_msprime_simplified_paths
  )
})

standard_msprime_tree_paths
```
