---
title: "test MaCSTS 1"
output: html_document
date: "2026-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## General Notes

- `MaCSTS()` now returns ts tables per chromosome: records ancestry from MaCS local trees (nodes/edges/intervals), and with `useMacsMut = TRUE` it also records MaCS-generated sites/mutations during simulation.
- TS ancestry/mutation work is split into staged helpers: `simAnc()`, `simMut()`, and `finalizeInbredTs()` (optional, when `inbred=TRUE`).
- `MaCSTS()`is designed to be a part of the high-level `runMacsTS()` wrapper, which will handle:

   - 1.TS generation (simAnc, and optionally simMut depending on mode);
   - 2.TS-to-founder conversion (logic in `R/makeFoundersFromTs.R`: site sampling/filtering, genMap construction);
   - 3.return a MapPop object like runMacs.
   
- Current TS chromosome count uses `nChr` (not `maxSites`, `maxSites` will only be in high level wrapper runMacsTS).
- Per-chromosome simulation runs in parallel when OpenMP is available (`nThreads`).

## Goal Of This Notebook

First validate `useMacsMut = TRUE`, because it should be directly comparable to legacy `MaCS` under the same args and seeds.

Note: when `useMacsMut = FALSE`, even ancestry (trees without mutation) can differ from MaCS because of RNG state changes! Specifically, it’s one shared RNG state/stream; MaCS (and MaCSTS with useMacsMut=TRUE) consume draws for both ancestry and mutation; MaCSTS with useMacsMut=FALSE skips mutation draws, so change later RNG state. 

## Helpers to Compare MaCSTS with MaCS

Helper overview:

- extract_macs_chr(): get site positions + haplotypes from legacy MaCS output.
- extract_ts_chr(): get site positions + haplotypes from TS tables via variant iterator.
- compare_chr(): run strict and order-robust chromosome-level equivalence checks.
- run_case(): run MaCS and MaCSTS(useMacsMut=TRUE) with matched args/seeds and compare.

```{r}
devtools::load_all()
library(RcppTskit)

# Convert raw 00/01 display values to integer 0/1 matrix.
to_int01 <- function(x) {
  matrix(as.integer(x), nrow = nrow(x), ncol = ncol(x), dimnames = dimnames(x))
}

# Read one chromosome from MaCS output (genMap + unpacked haplotypes).
extract_macs_chr <- function(macs_out, chr = 1L, nThreads = 1L) {
  pos <- as.numeric(macs_out$genMap[[chr]])
  n_sites <- length(pos)
  if (n_sites == 0L) {
    return(list(pos = numeric(0), hap = matrix(integer(0), nrow = 0, ncol = 0)))
  }
  chr_geno <- macs_out$geno[chr]
  hap_raw <- AlphaSimR:::getHaplo(
    geno = chr_geno,
    lociPerChr = as.integer(n_sites),
    lociLoc = as.integer(seq_len(n_sites)),
    nThreads = as.integer(nThreads)
  )
  list(pos = pos, hap = to_int01(hap_raw))
}

# Read one chromosome from TS tables (positions + variant genotypes).
extract_ts_chr <- function(tc_xptr) {
  tc <- RcppTskit::TableCollection$new(xptr = tc_xptr)
  ts <- tc$tree_sequence()
  n_samples <- as.integer(ts$num_samples())

  it <- ts$variants()
  pos <- numeric(0)
  cols <- list()
  repeat {
    v <- it$next_variant()
    if (is.null(v)) break
    pos <- c(pos, as.numeric(v$position))
    cols[[length(cols) + 1L]] <- as.integer(v$genotypes)
  }

  hap <- if (length(cols) == 0L) {
    matrix(integer(0), nrow = n_samples, ncol = 0L)
  } else {
    do.call(cbind, cols)
  }

  list(
    pos = pos,
    hap = hap,
    num_sites = as.integer(ts$num_sites()),
    num_mutations = as.integer(ts$num_mutations())
  )
}

# Sort position/genotype columns by position for strict positional matching.
sort_by_pos <- function(x) {
  if (length(x$pos) == 0L) return(x)
  o <- order(x$pos)
  list(pos = x$pos[o], hap = x$hap[, o, drop = FALSE])
}

# Build position+haplotype keys for multiset comparison.
site_hap_keys <- function(pos, hap) {
  if (length(pos) == 0L) return(character(0))
  p <- format(signif(pos, 15), scientific = FALSE, trim = TRUE)
  vapply(
    seq_len(ncol(hap)),
    function(j) paste0(p[j], "|", paste(hap[, j], collapse = "")),
    character(1)
  )
}

# Compare one chromosome:
# 1) strict check after sorting by position,
# 2) multiset check to tolerate column-order differences.
compare_chr <- function(macs_chr, ts_chr) {
  A <- sort_by_pos(macs_chr)
  B <- sort_by_pos(ts_chr)

  same_nsites <- ncol(A$hap) == ncol(B$hap)
  same_positions_strict <- isTRUE(all.equal(A$pos, B$pos, tolerance = 0))
  same_hap_strict <- identical(A$hap, B$hap)

  # Order-robust check for duplicate-position edge cases
  keysA <- site_hap_keys(macs_chr$pos, macs_chr$hap)
  keysB <- site_hap_keys(ts_chr$pos, ts_chr$hap)
  same_site_hap_multiset <- identical(sort(keysA), sort(keysB))

  data.frame(
    same_nsites = same_nsites,
    same_positions_strict = same_positions_strict,
    same_hap_strict = same_hap_strict,
    same_site_hap_multiset = same_site_hap_multiset,
    ts_num_sites = ts_chr$num_sites,
    ts_num_mutations = ts_chr$num_mutations
  )
}

# Run a full matched-seed comparison for nChr chromosomes.
run_case <- function(args, nChr, inbred, ploidy, seed, nThreads = 1L) {
  seed_vec <- rep(as.integer(seed), as.integer(nChr))

  macs <- AlphaSimR:::MaCS(
    args = args,
    maxSites = rep(0L, as.integer(nChr)),
    inbred = inbred,
    ploidy = as.integer(ploidy),
    nThreads = as.integer(nThreads),
    seed = seed_vec
  )

  ts_out <- AlphaSimR:::MaCSTS(
    args = args,
    nChr = as.integer(nChr),
    inbred = inbred,
    ploidy = as.integer(ploidy),
    nThreads = as.integer(nThreads),
    seed = seed_vec,
    usePhysicalPositions = FALSE,
    useMacsMut = TRUE
  )

  out <- lapply(seq_len(as.integer(nChr)), function(chr) {
    m <- extract_macs_chr(macs, chr = chr, nThreads = nThreads)
    t <- extract_ts_chr(ts_out$tables[[chr]])
    cbind(chr = chr, compare_chr(m, t))
  })
  do.call(rbind, out)
}
```

## General Case
In the first case, I will use some more detail results to show what are the objects being compared here:
```{r}
args <- "4 10000 -t 1e-3 -r 1e-2 -s "
seed <- as.integer(42)
inbred <- FALSE
ploidy <- 2L
nChr <- 1L

macs <- AlphaSimR:::MaCS(
    args = args,
    maxSites = rep(0L, as.integer(nChr)),
    inbred = inbred,
    ploidy = ploidy,
    seed = seed,
    nThreads = 1L
  )

  ts_out <- AlphaSimR:::MaCSTS(
    args = args,
    nChr = nChr,
    inbred = inbred,
    ploidy = ploidy,
    seed = seed,
    usePhysicalPositions = FALSE,
    useMacsMut = TRUE,
    nThreads = 1L
  )

chr = 1L  
m <- extract_macs_chr(macs, chr = chr)
t <- extract_ts_chr(ts_out$tables[[chr]])
```

Both genotypes and positions matched between MaCS and MaCSTS with `useMacsMut = TRUE`:

```
> print(m)
$pos
 [1] 0.1589878 0.1677192 0.1945866 0.2692034 0.2851194 0.2942718
 [7] 0.3008355 0.3116991 0.3385464 0.3423128 0.3543942 0.4430878
[13] 0.4690506 0.4930433 0.5781099 0.6904748 0.7921847 0.8604606

$hap
     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10] [,11] [,12]
[1,]    1    0    0    0    0    0    1    0    1     0     0     0
[2,]    1    0    1    1    1    1    0    1    0     0     0     0
[3,]    0    1    1    0    0    0    1    0    1     1     1     1
[4,]    0    1    1    0    0    0    1    0    0     0     0     0
     [,13] [,14] [,15] [,16] [,17] [,18]
[1,]     0     0     0     0     1     0
[2,]     0     0     1     0     1     0
[3,]     0     1     1     0     0     0
[4,]     1     0     0     1     0     1
```
```
> print(t)
$pos
 [1] 0.1589878 0.1677192 0.1945866 0.2692034 0.2851194 0.2942718
 [7] 0.3008355 0.3116991 0.3385464 0.3423128 0.3543942 0.4430878
[13] 0.4690506 0.4930433 0.5781099 0.6904748 0.7921847 0.8604606

$hap
     [,1] [,2] [,3] [,4] [,5] [,6] [,7] [,8] [,9] [,10] [,11] [,12]
[1,]    1    0    0    0    0    0    1    0    1     0     0     0
[2,]    1    0    1    1    1    1    0    1    0     0     0     0
[3,]    0    1    1    0    0    0    1    0    1     1     1     1
[4,]    0    1    1    0    0    0    1    0    0     0     0     0
     [,13] [,14] [,15] [,16] [,17] [,18]
[1,]     0     0     0     0     1     0
[2,]     0     0     1     0     1     0
[3,]     0     1     1     0     0     0
[4,]     1     0     0     1     0     1

$num_sites
[1] 18

$num_mutations
[1] 18
```

Or we can run:
```{r}
z <- run_case(
      args = args,
      nChr = 1L,
      inbred = inbred,
      ploidy = ploidy,
      seed = seed
    )
```
`same_hap_strict = TRUE` plus `same_positions_strict = TRUE` means direct compatibility matched exactly:
```
> print(z)
  chr same_nsites same_positions_strict same_hap_strict
1   1        TRUE                  TRUE            TRUE
  same_site_hap_multiset ts_num_sites ts_num_mutations
1                   TRUE           18               18
```

## More cases
Now I will run a scenario sweep with 5 different scenarios, and all of them show the same result that MaCSTS with `useMacsMut = TRUE` can reproduce the same sites and haplotypes as MaCS.

### Scenario1: ploidy = 1L
```{r}
args <- "4 10000 -t 1e-3 -r 1e-2 -s "
seed <- as.integer(42)
inbred <- FALSE
ploidy <- 1L
nChr <- 1L
z1 <- run_case(
      args = args,
      nChr = 1L,
      inbred = inbred,
      ploidy = ploidy,
      seed = seed
    )
print(z1)
```

### Scenario2: inbred = TRUE
```{r}
args <- "4 10000 -t 1e-3 -r 1e-2 -s "
seed <- as.integer(42)
inbred <- TRUE
ploidy <- 2L
nChr <- 1L
z2 <- run_case(
      args = args,
      nChr = 1L,
      inbred = inbred,
      ploidy = ploidy,
      seed = seed
    )
print(z2)
```

### Scenario3: demography
```{r}
# eN 0.2 2.0: at time 0.2, size jumps to 2x baseline.
# eN 1.0 0.5: at time 1.0, size jumps to 0.5x baseline.
args <- "4 10000 -t 1e-3 -r 1e-2 -eN 0.2 2.0 -eN 1.0 0.5 -s "
seed <- as.integer(42)
inbred <- FALSE
ploidy <- 2L
nChr <- 1L
z3 <- run_case(
      args = args,
      nChr = 1L,
      inbred = inbred,
      ploidy = ploidy,
      seed = seed
    )
print(z3)
```


### Scenario4: 2 populations after a split
```{r}
# -I: 8 sampled chromosomes from 2 pops total, with 4 from pop 1 and 4 from pop 2
# -ej: at time 1.0, pop 2 merges into pop 1 (backward), which corresponds to a split from pop 1 to pop 2 forward in time.
args <- "8 10000 -t 1e-3 -r 1e-2 -I 2 4 4 -ej 1.0 2 1 -s "
seed <- as.integer(42)
inbred <- FALSE
ploidy <- 2L
nChr <- 1L

ts_out4 <- AlphaSimR:::MaCSTS(
  args = args,
  nChr = nChr,
  inbred = inbred,
  ploidy = ploidy,
  seed = seed,
  usePhysicalPositions = FALSE,
  useMacsMut = TRUE,
  nThreads = 1L
)

# read population info from node table for the 4th node (which should be in pop 2 after the split)
tc4 <- RcppTskit::TableCollection$new(xptr = ts_out4$tables[[1]])
tc4$node_table_get_row(4)['population']
tc4$node_table_get_row(3)['population']

z4 <- run_case(
      args = args,
      nChr = 1L,
      inbred = inbred,
      ploidy = ploidy,
      seed = seed
    )
print(z4)
```

### Scenario5: 2 populations with a migration
```{r}
# -I: 8 sampled chromosomes from 2 pops total, with 4 from pop 1 and 4 from pop 2
# at backward time t = 0.5, set the global migration-rate parameter to 5e-3 for all populations
args <- "8 10000 -t 1e-3 -r 1e-2 -I 2 4 4 -eM 0.5 5e-3 -s "
seed <- as.integer(42)
inbred <- FALSE
ploidy <- 2L
nChr <- 1L

z5 <- run_case(
      args = args,
      nChr = 1L,
      inbred = inbred,
      ploidy = ploidy,
      seed = seed
    )
print(z5)
```


