---
title: "Forward TS recorder: tree input versus VCF input"
output: html_document
date: "2026-08-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

project_root <- normalizePath(if (file.exists("DESCRIPTION")) "." else "..")
local_lib <- file.path(project_root, ".Rlib")
if (dir.exists(local_lib)) {
  .libPaths(c(local_lib, .libPaths()))
}
library(AlphaSimR)
library(RcppTskit)
```

This is a short companion to `dev/testTsForwardRecorder.Rmd`. It uses the same
msprime forward-simulation parameters, but starts the founders in two ways:

1. directly from `msprime_chr0.trees` and `msprime_chr1.trees`;
2. from the matching phased VCF, `msprime_combined.vcf`, with `tsRecord=TRUE`.

The final `.trees` files are not expected to be byte-identical: the tree input
keeps the original ancestral history and source sites, while the VCF input
initializes synthetic founder nodes from haplotypes. The VCF importer also has
an `addTsMut` option. When `addTsMut=FALSE`, VCF marker sites are not written as
synthetic TS mutation events; when `addTsMut=TRUE`, retained VCF marker
positions are added as site rows and allele-1 calls are added as mutation rows
on founder haplotype nodes. Most of this notebook uses `addTsMut=FALSE`; the
last section compares both modes. The validation below checks the common part:
same retained founder haplotypes, same genetic map, same forward breeding loop,
and same forward-added inheritance edges.

## Helpers

These are the same validation helpers used in `testTsForwardRecorder.Rmd`,
trimmed to the pieces needed here.

```{r helpers}
normalize_edge_df <- function(df) {
  out <- data.frame(
    chr = as.integer(df$chr),
    child = as.integer(df$child),
    parent = as.integer(df$parent),
    left = as.numeric(df$left),
    right = as.numeric(df$right),
    stringsAsFactors = FALSE
  )
  out <- out[out$right > out$left, , drop = FALSE]
  if (nrow(out) == 0L) {
    return(out)
  }
  out <- out[order(out$chr, out$child, out$parent, out$left, out$right), , drop = FALSE]
  rownames(out) <- NULL
  out
}

collect_forward_edges_for_children <- function(recorder, chr, child_iids) {
  chrState <- recorder$chr[[chr]]
  tc <- chrState$tc
  n_edge <- as.integer(tc$num_edges())

  iid_keys <- ls(chrState$indMap, all.names = TRUE)
  if (length(iid_keys) == 0L || n_edge == 0L) {
    return(data.frame(chr = integer(), child = integer(), parent = integer(),
                      left = numeric(), right = numeric(), stringsAsFactors = FALSE))
  }

  ind_rows <- vapply(iid_keys, function(key) {
    as.integer(get(key, envir = chrState$indMap, inherits = FALSE))
  }, integer(1))
  ind_row_to_iid <- setNames(as.integer(iid_keys), as.character(ind_rows))

  out <- vector("list", n_edge)
  k <- 0L
  for (i in seq_len(n_edge)) {
    e <- tc$edge_table_get_row(i - 1L)
    child_ind_row <- as.integer(tc$node_table_get_row(as.integer(e$child))$individual)
    parent_ind_row <- as.integer(tc$node_table_get_row(as.integer(e$parent))$individual)
    child_iid <- ind_row_to_iid[as.character(child_ind_row)]
    parent_iid <- ind_row_to_iid[as.character(parent_ind_row)]

    if (length(child_iid) == 0L || length(parent_iid) == 0L ||
        is.na(child_iid) || is.na(parent_iid)) {
      next
    }
    child_iid <- as.integer(child_iid)
    if (!(child_iid %in% child_iids)) {
      next
    }

    k <- k + 1L
    out[[k]] <- data.frame(
      chr = as.integer(chr),
      child = child_iid,
      parent = as.integer(parent_iid),
      left = as.numeric(e$left),
      right = as.numeric(e$right),
      stringsAsFactors = FALSE
    )
  }

  if (k == 0L) {
    return(data.frame(chr = integer(), child = integer(), parent = integer(),
                      left = numeric(), right = numeric(), stringsAsFactors = FALSE))
  }
  do.call(rbind, out[seq_len(k)])
}

collect_forward_edges_all_chr <- function(recorder, child_iids) {
  out <- lapply(seq_len(recorder$nChr), function(chr) {
    collect_forward_edges_for_children(recorder, chr = chr, child_iids = child_iids)
  })
  normalize_edge_df(do.call(rbind, out))
}

same_gen_map <- function(a, b, tol = 1e-12) {
  length(a) == length(b) &&
    all(mapply(function(x, y) {
      isTRUE(all.equal(as.numeric(x), as.numeric(y), tolerance = tol))
    }, a, b))
}

edge_example_table <- function(tree_edges, vcf_edges, n = 8L) {
  n <- min(n, nrow(tree_edges), nrow(vcf_edges))
  if (n == 0L) {
    return(data.frame())
  }
  data.frame(
    row = seq_len(n),
    chr = tree_edges$chr[seq_len(n)],
    child = tree_edges$child[seq_len(n)],
    parent = tree_edges$parent[seq_len(n)],
    tree_left = tree_edges$left[seq_len(n)],
    vcf_left = vcf_edges$left[seq_len(n)],
    left_diff = tree_edges$left[seq_len(n)] - vcf_edges$left[seq_len(n)],
    tree_right = tree_edges$right[seq_len(n)],
    vcf_right = vcf_edges$right[seq_len(n)],
    right_diff = tree_edges$right[seq_len(n)] - vcf_edges$right[seq_len(n)],
    stringsAsFactors = FALSE
  )
}

edge_example_table_by_chr <- function(tree_edges, vcf_edges, n = 5L) {
  out <- lapply(split(seq_len(nrow(tree_edges)), tree_edges$chr), function(idx) {
    chr <- tree_edges$chr[idx[[1L]]]
    vcf_idx <- which(vcf_edges$chr == chr)
    edge_example_table(
      tree_edges = tree_edges[idx, , drop = FALSE],
      vcf_edges = vcf_edges[vcf_idx, , drop = FALSE],
      n = n
    )
  })
  do.call(rbind, out)
}

run_standard_forward <- function(founder, seed = 42L) {
  old_opts <- options(
    AlphaSimR.tsForwardKeepSeg = FALSE,
    AlphaSimR.tsForwardKeepRecHistGen = TRUE
  )
  on.exit(options(old_opts), add = TRUE)

  set.seed(seed)
  SP <- SimParam$new(founder)
  SP$nThreads <- 1L
  SP$quadProb <- 0
  SP$setSexes("no")
  SP$addTraitA(nQtlPerChr = 5, mean = 500, var = 450)
  SP$setTrackPed(TRUE)
  SP$setTrackRecGen(TRUE)

  basePop <- newPop(founder, simParam = SP)
  SP$setTrackTs(TRUE, founderPop = basePop)
  basePop <- setPheno(basePop, h2 = 0.5, simParam = SP)

  sim_output <- list(basePop)
  for (cycle in seq_len(2L)) {
    prev <- sim_output[[cycle]]
    chosen <- selectInd(pop = prev, nInd = min(2L, nInd(prev)), use = "gv", simParam = SP)
    offspring <- randCross(pop = chosen, nCrosses = 1L, nProgeny = 5L, simParam = SP)
    offspring <- setPheno(pop = offspring, h2 = 0.5, simParam = SP)
    sim_output[[cycle + 1L]] <- offspring
  }

  child_iids <- unlist(lapply(sim_output[-1L], function(pop) as.integer(pop@iid)), use.names = FALSE)
  recorder <- attr(SP, "tsForwardRecorder", exact = TRUE)
  edges <- collect_forward_edges_all_chr(recorder, child_iids = child_iids)

  list(SP = SP, sim_output = sim_output, recorder = recorder, edges = edges)
}
```

## Founder Imports

The recombination map, `segSites`, and site-sampling seed match the msprime
section of `testTsForwardRecorder.Rmd`.

```{r imports}
out_dir <- file.path(project_root, "dev", "testData")
L1 <- 1e6
L2 <- 2e6

chr_info <- list(
  list(
    ts_path = file.path(out_dir, "msprime_chr0.trees"),
    breaks = c(0, L1 / 2, L1),
    rates = c(5e-5, 5e-6),
    segSites = 60L
  ),
  list(
    ts_path = file.path(out_dir, "msprime_chr1.trees"),
    breaks = c(0, L2 / 3, 2 * L2 / 3, L2),
    rates = c(5e-7, 5e-6, 5e-7),
    segSites = 155L
  )
)

stopifnot(all(file.exists(vapply(chr_info, `[[`, character(1), "ts_path"))))
stopifnot(file.exists(file.path(out_dir, "msprime_combined.vcf")))

tree_founder <- AlphaSimR:::asMapPop(
  chr_info = chr_info,
  inbred = FALSE,
  ploidy = 2L,
  site_sampling_seed = 42L,
  returnMeta = TRUE
)

vcf_founder <- importVCF(
  vcfFile = file.path(out_dir, "msprime_combined.vcf"),
  breaks = list(chr0 = chr_info[[1L]]$breaks, chr1 = chr_info[[2L]]$breaks),
  rates = list(chr0 = chr_info[[1L]]$rates, chr1 = chr_info[[2L]]$rates),
  segSites = c(chr0 = 60L, chr1 = 155L),
  siteSamplingSeed = 42L,
  missing = "error",
  tsRecord = TRUE,
  addTsMut = FALSE,
  returnMeta = TRUE
)
```

## Validate Founder Equivalence

```{r founder-equivalence}
tree_hap <- pullSegSiteHaplo(tree_founder$pop, nThreads = 1L)
vcf_hap <- pullSegSiteHaplo(vcf_founder$pop, nThreads = 1L)
vcf_ts_summary <- lapply(vcf_founder$tsTables, rtsk_table_collection_summary2)

founder_checks <- data.frame(
  check = c(
    "nInd",
    "ploidy",
    "nLoci",
    "retained physical positions",
    "genetic map numeric values",
    "haplotype numeric values",
    "VCF has TS forward source",
    "VCF TS site tables are empty",
    "VCF TS mutation tables are empty"
  ),
  passed = c(
    tree_founder$pop@nInd == vcf_founder$pop@nInd,
    tree_founder$pop@ploidy == vcf_founder$pop@ploidy,
    identical(tree_founder$pop@nLoci, vcf_founder$pop@nLoci),
    identical(unname(tree_founder$keptPosBp), unname(vcf_founder$keptPos)),
    same_gen_map(tree_founder$pop@genMap, vcf_founder$pop@genMap),
    identical(unname(tree_hap), unname(vcf_hap)),
    !is.null(attr(vcf_founder$pop, "tsForwardSource", exact = TRUE)),
    all(vapply(vcf_ts_summary, function(x) as.integer(x$num_sites) == 0L, logical(1))),
    all(vapply(vcf_ts_summary, function(x) as.integer(x$num_mutations) == 0L, logical(1)))
  )
)
knitr::kable(founder_checks)
stopifnot(all(founder_checks$passed))
```

## Run The Same Forward Simulation

```{r forward}
tree_run <- run_standard_forward(tree_founder$pop, seed = 42L)
vcf_run <- run_standard_forward(vcf_founder$pop, seed = 42L)
```

## Validate Forward Recorder Equivalence

```{r forward-equivalence}
tree_edges <- tree_run$edges
vcf_edges <- vcf_run$edges
n <- min(nrow(tree_edges), nrow(vcf_edges))

edge_checks <- data.frame(
  check = c(
    "forward edge rows",
    "chr/child/parent ordering",
    "left coordinates",
    "right coordinates"
  ),
  passed = c(
    nrow(tree_edges) == nrow(vcf_edges),
    identical(tree_edges[, c("chr", "child", "parent"), drop = FALSE],
              vcf_edges[, c("chr", "child", "parent"), drop = FALSE]),
    isTRUE(all.equal(tree_edges$left, vcf_edges$left, tolerance = 1e-6)),
    isTRUE(all.equal(tree_edges$right, vcf_edges$right, tolerance = 1e-6))
  )
)
knitr::kable(edge_checks)
stopifnot(all(edge_checks$passed))

edge_summary <- data.frame(
  source = c("trees", "vcf"),
  edge_rows = c(nrow(tree_edges), nrow(vcf_edges)),
  max_left_diff = if (n == 0L) NA_real_ else max(abs(tree_edges$left[seq_len(n)] - vcf_edges$left[seq_len(n)])),
  max_right_diff = if (n == 0L) NA_real_ else max(abs(tree_edges$right[seq_len(n)] - vcf_edges$right[seq_len(n)]))
)
knitr::kable(edge_summary, digits = 6)

knitr::kable(
  edge_example_table_by_chr(tree_edges, vcf_edges, n = 5L),
  digits = 3,
  format.args = list(scientific = FALSE, nsmall = 3)
)
```

## Write Forward TS Files

```{r write-trees}
tree_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
  tree_run$SP,
  out_dir = out_dir,
  out_basename = "standard_msprime_tree_input_forward_recorder",
  clear = TRUE
)

vcf_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
  vcf_run$SP,
  out_dir = out_dir,
  out_basename = "standard_msprime_vcf_input_forward_recorder",
  clear = TRUE
)

tree_files <- data.frame(
  chr = seq_along(tree_paths),
  tree_input_path = tree_paths,
  vcf_input_path = vcf_paths,
  tree_exists = file.exists(tree_paths),
  vcf_exists = file.exists(vcf_paths),
  stringsAsFactors = FALSE
)
knitr::kable(tree_files)
stopifnot(all(tree_files$tree_exists), all(tree_files$vcf_exists))
```

## Missing-Data VCF Smoke Test

This uses the same two-chromosome VCF with missing genotypes. Missing sites are
filtered before reservoir sampling. This is not expected to match the complete
tree input at all loci, but it should still initialize founder nodes, keep TS
site/mutation tables empty with `addTsMut=FALSE`, and record the forward
simulation.

```{r missing-vcf}
missing_vcf <- file.path(out_dir, "msprime_combined_missing.vcf")
stopifnot(file.exists(missing_vcf))

missing_founder <- importVCF(
  vcfFile = missing_vcf,
  breaks = list(chr0 = chr_info[[1L]]$breaks, chr1 = chr_info[[2L]]$breaks),
  rates = list(chr0 = chr_info[[1L]]$rates, chr1 = chr_info[[2L]]$rates),
  segSites = c(chr0 = 40L, chr1 = 100L),
  siteSamplingSeed = 42L,
  missing = "filter",
  tsRecord = TRUE,
  addTsMut = FALSE,
  returnMeta = TRUE
)

missing_run <- run_standard_forward(missing_founder$pop, seed = 42L)
missing_paths <- AlphaSimR:::tsForwardFinalizeFromSimParam(
  missing_run$SP,
  out_dir = out_dir,
  out_basename = "standard_msprime_vcf_missing_input_forward_recorder",
  clear = TRUE
)

missing_summary <- data.frame(
  nInd = missing_founder$pop@nInd,
  ploidy = missing_founder$pop@ploidy,
  nLoci = paste(missing_founder$pop@nLoci, collapse = ","),
  kept = missing_founder$stats$kept,
  skippedMissing = missing_founder$stats$skippedMissing,
  forwardEdgeRows = nrow(missing_run$edges),
  treeFilesWritten = all(file.exists(missing_paths)),
  stringsAsFactors = FALSE
)
knitr::kable(missing_summary)
stopifnot(missing_summary$treeFilesWritten)
```

## Python Inspection Of Generated Tree Files

This mirrors the Python/tskit inspection section in `testTsForwardRecorder.Rmd`,
but points at the newly generated tree-input, VCF-input, and missing-VCF-input
files from this notebook.

```{r py-setup}
# Use your own path here if needed.
reticulate::use_python("/usr/bin/python3", required = TRUE)
library(reticulate)
reticulate::py_config()
```

```{r python-paths}
python_files <- data.frame(
  label = c(
    paste0("tree input chr", seq_along(tree_paths) - 1L),
    paste0("vcf input chr", seq_along(vcf_paths) - 1L),
    paste0("vcf missing input chr", seq_along(missing_paths) - 1L)
  ),
  source = rep(c("tree_input", "vcf_input", "vcf_missing_input"),
               c(length(tree_paths), length(vcf_paths), length(missing_paths))),
  chr = c(seq_along(tree_paths), seq_along(vcf_paths), seq_along(missing_paths)) - 1L,
  path = normalizePath(c(tree_paths, vcf_paths, missing_paths), mustWork = FALSE),
  stringsAsFactors = FALSE
)
knitr::kable(python_files)

library(reticulate)
files_obj <- lapply(seq_len(nrow(python_files)), function(i) {
  as.list(python_files[i, , drop = FALSE])
})
reticulate::py_run_string("generated_files = r.files_obj")
```

```{python compare-generated-tree-files}
from pathlib import Path
import tskit

for row in generated_files:
    ts_path = Path(row["path"])
    ts = tskit.load(ts_path)
    print(row["label"])
    print("  path:", ts_path)
    print("  nodes:", ts.num_nodes)
    print("  edges:", ts.num_edges)
    print("  trees:", ts.num_trees)
    print("  sites:", ts.num_sites)
    print("  mutations:", ts.num_mutations)
```

Print the first tree for chromosome 0 from each generated input path.

```{python print-first-tree}
from pathlib import Path
import tskit

for row in generated_files:
    if int(row["chr"]) != 0:
        continue
    ts_path = Path(row["path"])
    ts = tskit.load(ts_path)

    print(f"\n{row['label']}")
    print("  path:", ts_path)
    print("  trees:", ts.num_trees)
    print("  nodes:", ts.num_nodes)
    print("  edges:", ts.num_edges)
    print(ts.first().draw_text())
```

Print one corresponding non-missing chr0 node from the tree-input and VCF-input
outputs. The numeric node IDs differ because the tree-input output keeps the
original source ancestry and source nodes, while the VCF-input output starts
from synthetic founder nodes only. The AlphaSimR node metadata identifies the
same logical node.

```{python print-specific-corresponding-node}
from pathlib import Path
import json
import tskit

def generated_path(source, chr_index):
    for row in generated_files:
        if row["source"] == source and int(row["chr"]) == chr_index:
            return Path(row["path"])
    raise ValueError(f"No generated file for source={source}, chr={chr_index}")

def alpha_node_id(node):
    meta = node.metadata
    if isinstance(meta, bytes):
        meta = json.loads(meta.decode("utf-8")) if meta else {}
    return meta.get("alphaSimR", {}).get("id")

tree_chr0 = tskit.load(generated_path("tree_input", 0))
vcf_chr0 = tskit.load(generated_path("vcf_input", 0))

tree_node = tree_chr0.node(260)
vcf_node = vcf_chr0.node(4)

print(tree_node)
print("individual:", tree_chr0.individual(tree_node.individual))
print(vcf_node)
print("individual:", vcf_chr0.individual(vcf_node.individual))

print("  same AlphaSimR node id:", alpha_node_id(tree_node) == alpha_node_id(vcf_node))
print("  AlphaSimR node id:", alpha_node_id(tree_node))
print("  same flags:", tree_node.flags == vcf_node.flags)
print("  same time:", tree_node.time == vcf_node.time)
print("  same population:", tree_node.population == vcf_node.population)
```

Print and draw pedigrees from the generated files, following the same style as
`testTsForwardRecorder.Rmd`.

```{python print-pedigrees}
from pathlib import Path
import tskit
from matplotlib import pyplot as plt
import networkx as nx

def draw_pedigree(ped_ts, title):
    plt.figure(figsize=(8, 5))

    G = nx.DiGraph()
    for ind in ped_ts.individuals():
        if len(ind.nodes) == 0:
            continue
        time = ped_ts.node(ind.nodes[0]).time
        pop = ped_ts.node(ind.nodes[0]).population
        G.add_node(ind.id, time=time, population=pop)
        for p in ind.parents:
            if p != tskit.NULL:
                G.add_edge(ind.id, p)

    pos = nx.multipartite_layout(G, subset_key="time", align="horizontal")
    colours = plt.rcParams["axes.prop_cycle"].by_key()["color"]
    node_colours = [
        colours[attr["population"] % len(colours)]
        for attr in G.nodes.values()
    ]

    nx.draw_networkx(G, pos, with_labels=True, node_color=node_colours)
    plt.title(title)
    plt.show()
    plt.close()

print(generated_files)
for row in generated_files:
    if row["label"] not in ["tree input chr1", "vcf input chr1", "vcf missing input chr1"]:
        continue

    print(row)
    ts = tskit.load(row["path"])
    print(f"\n{row['label']}")
    draw_pedigree(ts, row["label"])
```

## Effect Of `addTsMut`

For VCF input, `addTsMut` only controls whether retained VCF marker states are
materialized into the initial TS site/mutation tables. It should not change the
forward recorder's pedigree, nodes, edges, or tree intervals. With
`addTsMut=TRUE`, the output gains:

- one site row per retained VCF marker position;
- one mutation row for each founder haplotype carrying allele `1` at that site.

```{r addTsMut-comparison}
vcf_founder_add_mut <- importVCF(
  vcfFile = file.path(out_dir, "msprime_combined.vcf"),
  breaks = list(chr0 = chr_info[[1L]]$breaks, chr1 = chr_info[[2L]]$breaks),
  rates = list(chr0 = chr_info[[1L]]$rates, chr1 = chr_info[[2L]]$rates),
  segSites = c(chr0 = 60L, chr1 = 155L),
  siteSamplingSeed = 42L,
  missing = "error",
  tsRecord = TRUE,
  addTsMut = TRUE,
  returnMeta = TRUE
)

vcf_run_add_mut <- run_standard_forward(vcf_founder_add_mut$pop, seed = 42L)
vcf_paths_add_mut <- AlphaSimR:::tsForwardFinalizeFromSimParam(
  vcf_run_add_mut$SP,
  out_dir = out_dir,
  out_basename = "standard_msprime_vcf_addTsMut_forward_recorder",
  clear = TRUE
)

tskit <- reticulate::import("tskit")
ts_counts <- function(paths, addTsMut) {
  do.call(rbind, lapply(seq_along(paths), function(i) {
    ts <- tskit$load(paths[[i]])
    data.frame(
      addTsMut = addTsMut,
      chr = i - 1L,
      nodes = as.integer(ts$num_nodes),
      edges = as.integer(ts$num_edges),
      trees = as.integer(ts$num_trees),
      sites = as.integer(ts$num_sites),
      mutations = as.integer(ts$num_mutations),
      stringsAsFactors = FALSE
    )
  }))
}

addTsMut_counts <- rbind(
  ts_counts(vcf_paths, addTsMut = FALSE),
  ts_counts(vcf_paths_add_mut, addTsMut = TRUE)
)
knitr::kable(addTsMut_counts)

addTsMut_wide <- merge(
  addTsMut_counts[addTsMut_counts$addTsMut == FALSE, ],
  addTsMut_counts[addTsMut_counts$addTsMut == TRUE, ],
  by = "chr",
  suffixes = c("_false", "_true")
)
addTsMut_checks <- data.frame(
  chr = addTsMut_wide$chr,
  same_nodes = addTsMut_wide$nodes_false == addTsMut_wide$nodes_true,
  same_edges = addTsMut_wide$edges_false == addTsMut_wide$edges_true,
  same_trees = addTsMut_wide$trees_false == addTsMut_wide$trees_true,
  sites_added = addTsMut_wide$sites_false == 0L & addTsMut_wide$sites_true > 0L,
  mutations_added = addTsMut_wide$mutations_false == 0L & addTsMut_wide$mutations_true > 0L,
  stringsAsFactors = FALSE
)
knitr::kable(addTsMut_checks)
stopifnot(all(addTsMut_checks[, -1L]))
```
