---
title: "test MaCSTS 4: wrapper/staged workflow overview"
output: html_document
date: "2026-06-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## General Notes

This notebook summarizes the current TS workflow design and parameter effects,
based on `testMaCSTS1.Rmd`, `testMaCSTS2.Rmd`, and `testMaCSTS3.Rmd`.

It starts from the two supported ways to run:

1. high-level wrapper: `runMacsTS(...)`
2. staged low-level path (for developers and advanced users): `simAnc(...) -> simMut(...) -> finalizeInbredTs(...) -> asMapPop(...)`

## 1) Two Ways To Run

### A. High-level wrapper (`runMacsTS`)

`runMacsTS` is the default user-facing entry point. It is designed to feel like
`runMacs/runMacs2` but with TS-aware internals and mode control.

- Builds MaCS command from species/manual inputs
- Runs ancestry and mutation according to `mutationMode`
- Converts to `MapPop` (including optional `segSites` sampling)
- Optionally returns TS tables/metadata

### B. Staged workflow (`simAnc` + `simMut` + ...)

This path is for development/debugging and explicit control over each stage.

- `simAnc`: ancestry-only TS from MaCS engine (`useMacsMut=FALSE`)
- `simMut`: post-TS mutation placement on edge table (if `mutationMode="postTs"`)
- `finalizeInbredTs`: optional inbred leaf expansion in TS (if `inbred=TRUE` and `ploidy>1`)
- `asMapPop`: convert TS to `MapPop` (sampling/filtering + map construction)

## 2) Function Relationship

```{r relationship-diagram, results='asis'}
cat("
runMacsTS()
  ├─ command builder (species/manual)
  ├─ mutationMode = 'macs'  -> MaCSTS(..., useMacsMut=TRUE, expandInbredSamples=FALSE)
  ├─ mutationMode = 'postTs'-> simAnc(...) -> simMut(...)
  ├─ mutationMode = 'none'  -> simAnc(...) only (TS-only unless user mutates later)
  ├─ optional finalizeInbredTs(...)
  └─ asMapPop(...) -> MapPop

Lower layers:
  simAnc()  -> MaCSTS(... useMacsMut=FALSE ...)
  simMut()  -> tsMutateTableCollection(...) per chromosome
  finalizeInbredTs() -> tsFinalizeInbredTableCollection(...) per chromosome
  MaCSTS()  -> C++ simulator + TS recorder
")
```

## 3) Core Parameters And Their Effects

```{r param-table}
param_effects <- data.frame(
  parameter = c(
    "mutationMode",
    "useMacsMut (MaCSTS)",
    "usePhysicalPositions",
    "Nref",
    "segSites",
    "inbred + ploidy",
    "expandInbredTs",
    "seed",
    "mutSeed",
    "nThreads"
  ),
  where_used = c(
    "runMacsTS",
    "MaCSTS",
    "MaCSTS/asMapPop",
    "MaCSTS (timeScale)",
    "asMapPop",
    "MaCSTS/asMapPop/finalizeInbredTs",
    "runMacsTS/finalizeInbredTs",
    "simAnc/MaCSTS ancestry RNG",
    "simMut post-TS RNG",
    "MaCS chromosomes + asMapPop worker"
  ),
  main_effect = c(
    "Selects mutation path: macs vs postTs vs none",
    "Adds MaCS-style mutation during ancestry if TRUE",
    "TS coordinates in bp (TRUE) or [0,1] (FALSE)",
    "Rescales node/mutation times; sets time_units to generations",
    "Caps number of retained variants per chromosome in conversion",
    "Controls sample interpretation and output haplotype structure",
    "If TRUE with inbred/ploidy>1, duplicates TS sample leaves",
    "Determines ancestry reproducibility",
    "Determines post-TS mutation reproducibility",
    "Parallelism level when available"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(param_effects)
```

## Helpers to Compare Two Ways To Run

```{r comparison-helpers}
tc_counts <- function(tc_xptr) {
  tc <- RcppTskit::TableCollection$new(xptr = tc_xptr)
  data.frame(
    num_nodes = as.integer(tc$num_nodes()),
    num_edges = as.integer(tc$num_edges()),
    num_sites = as.integer(tc$num_sites()),
    num_mutations = as.integer(tc$num_mutations())
  )
}

collect_ts_counts <- function(tables, method) {
  do.call(rbind, lapply(seq_along(tables), function(chr) {
    cbind(chr = chr, method = method, tc_counts(tables[[chr]]))
  }))
}

compare_wrapper_vs_staged <- function(out, staged) {
  if (!requireNamespace("RcppTskit", quietly = TRUE)) {
    stop("RcppTskit is required for TS checks.")
  }
  if (is.null(out$pop) || is.null(out$tables)) {
    stop("out must contain both $pop and $tables")
  }
  if (is.null(staged$pop) || is.null(staged$tables)) {
    stop("staged must contain both $pop and $tables")
  }
  
  ts_counts_wrap <- collect_ts_counts(out$tables, method = "wrapper")
  ts_counts_stage <- collect_ts_counts(staged$tables, method = "staged")
  ts_counts <- rbind(ts_counts_wrap, ts_counts_stage)
  
  nChr <- length(out$pop@geno)
  same_geno <- vapply(seq_len(nChr), function(chr) {
    identical(out$pop@geno[[chr]], staged$pop@geno[[chr]])
  }, logical(1))
  same_genMap <- vapply(seq_len(nChr), function(chr) {
    isTRUE(all.equal(out$pop@genMap[[chr]], staged$pop@genMap[[chr]], tolerance = 0))
  }, logical(1))
  same_ts_counts <- identical(
    ts_counts_wrap[, c("num_nodes", "num_edges", "num_sites", "num_mutations")],
    ts_counts_stage[, c("num_nodes", "num_edges", "num_sites", "num_mutations")]
  )
  
  checks <- data.frame(
    metric = c("nLoci", "geno_all_chr", "genMap_all_chr", "ts_table_counts"),
    equal = c(
      identical(out$pop@nLoci, staged$pop@nLoci),
      all(same_geno),
      all(same_genMap),
      same_ts_counts
    ),
    stringsAsFactors = FALSE
  )
  
  list(ts_counts = ts_counts, checks = checks)
}

print_wrapper_vs_staged <- function(cmp) {
  print(cmp$ts_counts)
  print(cmp$checks)
  invisible(cmp)
}
```

## Minimal Usage Examples

### A. High-level wrapper mode

```{r ex-runMacsTS}
library(AlphaSimR)
devtools::load_all()
set.seed(1)
out <- AlphaSimR:::runMacsTS(
  nInd = 4,
  nChr = 2,
  segSites = 60,
  inbred = FALSE,
  ploidy = 2L,
  species = "GENERIC",
  mutationMode = "postTs",      # or "macs" / "none"
  usePhysicalPositions = FALSE,
  nThreads = 1L,
  returnTs = TRUE
)

# runMacsTS returns:
# out$pop      -> MapPop
# out$tables   -> TS table collections (if returnTs=TRUE)
```

### B. Staged mode

```{r ex-staged}
library(AlphaSimR)

# use the same args and seeds as high-level wrapper mode
args <- out$args
nChr <- 2L
seed_chr <- out$seed

anc <- AlphaSimR:::simAnc(
  args = args,
  nChr = nChr,
  inbred = FALSE,
  ploidy = 2L,
  nThreads = 1L,
  seed = seed_chr,
  usePhysicalPositions = FALSE,
  Nref = NA_real_
)

mut <- AlphaSimR:::simMut(anc)  # uses anc$dTheta and derived mutation seeds by default

# Optional for inbred/ploidy expansion in TS:
# mut <- AlphaSimR:::finalizeInbredTs(mut, inbred = TRUE, ploidy = 2L)

pop <- AlphaSimR:::asMapPop(
  chr_info = list(
    tables = mut$tables,
    breaks = list(c(0, 1)),
    rates = list(c(1))
  ),
  ploidy = 2L,
  inbred = FALSE,
  segSites = 60
)
```

## Validation

### Wrapper vs staged equivalence

```{r wrapper-vs-staged-equivalence}
# Reuse objects from the previous ection:
# - out  : runMacsTS(...) result
# - mut  : simMut(...) result from staged path
# - pop  : asMapPop(...) result from staged path

stopifnot(exists("out"), exists("mut"), exists("pop"))
cmp <- compare_wrapper_vs_staged(
  out = out,
  staged = list(
    tables = mut$tables,
    pop = pop
  )
)
print_wrapper_vs_staged(cmp)
```

### Wrapper vs staged (inbred=TRUE, ploidy>1)

```{r wrapper-vs-staged-inbred}
# Example: inbred TRUE with ploidy 2
set.seed(1)
out_inbred <- AlphaSimR:::runMacsTS(
  nInd = 4,
  nChr = 2,
  segSites = 60,
  inbred = TRUE,
  ploidy = 2L,
  species = "GENERIC",
  mutationMode = "postTs",
  usePhysicalPositions = TRUE,
  expandInbredTs = TRUE,
  nThreads = 1L,
  returnTs = TRUE,
  Nref = 10000
)

anc_inbred <- AlphaSimR:::simAnc(
  args = out_inbred$args,
  nChr = 2L,
  inbred = TRUE,
  ploidy = 2L,
  nThreads = 1L,
  seed = out_inbred$seed,
  usePhysicalPositions = TRUE,
  #Nref = NA_real_,
  Nref = 10000
)
mut_inbred <- AlphaSimR:::simMut(anc_inbred)
fin_inbred <- AlphaSimR:::finalizeInbredTs(mut_inbred, inbred = TRUE, ploidy = 2L)

pop_inbred <- AlphaSimR:::asMapPop(
  chr_info = list(
    tables = fin_inbred$tables,
    breaks = rep(list(c(0, 1)), 2L),
    rates = rep(list(c(1)), 2L)
  ),
  ploidy = 2L,
  inbred = TRUE,
  segSites = 60,
  site_sampling_seed = 42L,
  nThreads = 1L
)

cmp_inbred <- compare_wrapper_vs_staged(
  out = out_inbred,
  staged = list(
    tables = fin_inbred$tables,
    pop = pop_inbred
  )
)
print_wrapper_vs_staged(cmp_inbred)
```

Output a ts as an example and check it in Python:
```{r export-inbred-ts}
tables_src <- if (exists("fin_inbred")) {
  fin_inbred$tables
} else if (exists("out_inbred")) {
  out_inbred$tables
} else {
  stop("Need fin_inbred or out_inbred in environment. Run chunk 5.1b first.")
}
# chr1
tc_xptr <- tables_src[[1]]
tc <- RcppTskit::TableCollection$new(xptr = tc_xptr)
ts <- tc$tree_sequence()
ts$write('testData/inbred_test.trees')
```

```{r py-setup}
library(reticulate)
# Use your own path here!
use_python('/usr/bin/python3', required = TRUE)
```


```{python}
import numpy as np
import tskit
ts_inbred = tskit.load('testData/inbred_test.trees')
ts_inbred.samples()

print(ts_inbred.first().draw_text())
from IPython.display import display
svg_size = (800, 250)
svg_string = ts_inbred.first().draw_svg(
    size=svg_size, time_scale="rank", omit_sites=True,
    y_axis=True, y_label=" "  # optional: show a time scale on the left
)
with open("tmp_tree.svg", "w") as f:
    f.write(svg_string)
print(ts_inbred.tables.nodes[2612:2620])
print(ts_inbred.tables.nodes[0:4])

g_matrix = ts_inbred.genotype_matrix()
for hap in range(0, len(g_matrix[0]), 2):
    print(hap, hap+1)
    print(np.array_equal(g_matrix[:, hap], g_matrix[:, hap+1]))

```
Sampled nodes are duplicated as expected for inbred/ploidy expansion, from nodes 0-3. Each pair of hap (from the same individual) have same alleles.


### Parameter sensitivity checks
In tests/testthat/test-runMacsTS-sensitivity.R
